An exposition on the propriety of restricted Boltzmann machines

### 2016-09-03 writing/draft.Rmd
attempt to make degeneracy section make some sense

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} 
\end{align}

Let the normalizing constant in equation (\ref{eqn:pmf}) be denoted $\gamma(\boldsymbol \theta)$.It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.

### Near-degeneracy
In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}. In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.

The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, for the RBM is defined as
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is unfeasible for a model of any real size.

>>>>

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} 
\end{align}

Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) makes three kinds of potential model impropriety of concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationship to one another.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes. An exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distibrution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$. Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.

The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$ have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{S}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is infeasible for a model of any real size.
